---
title: "Cohesion Index Calculation"
author: "Simon Garnier & Johann Delcourt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    css: custom.css
    fig_caption: true
    fig_width: 9
    fig_height: 6
    number_sections: true
    theme: cosmo
    toc: true
csl: http://www.zotero.org/styles/plos-computational-biology
bibliography: library.bib
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
      equationNumbers: { 
            autoNumber: "all",
            formatNumber: function (n) {return n}
      } 
  }
});
</script>

--- 

# Introduction

This note explains the steps to calculate a new index that estimates the 
cohesion of a group of individuals in a radial arm maze (note that this index 
can also be used for all types of mazes or arenas divided into distinct zones). 
This index is an alternative to the classical methods used to measure group 
cohesion based on topological and metric relationships between individuals (see 
@Delcourt2012-dq). These methods cannot be used reliably with a radial arm maze
because individuals located in adjacent arms might be spatially close, yet 
incapable of interacting directly with each other.

The new cohesion index $I_c$ introduced here measures the ability of individuals
to form cohesive groups in a radial arm maze. Its value is maximal (i.e. $I_c=1$)
when all the animals are located in a single zone of the maze; it is minimal 
(i.e. $I_c=0$) when the animals are distributed uniformly between all the zones 
of the maze. 

[Back to top]("#")

---

# Definition of $I_c$

Let $f_i$ be the number of individuals in each zone $z_i$ of the radial arm maze. 
For a maze with $Z$ zones, the Euclidean distance $D_c$ between all $f_i$ can be 
calculated as:

$$ D_c = \sqrt{\sum_{i=1}^{Z}(f_i)^2} $$

Let $D_{min}$ be the value of $D_c$ for the least cohesive distribution possible
of individuals between the zones of the maze (i.e. homogenous distribution of 
the individuals), and $N$ the total number of individuals in the maze. Therefore 
the cohesion index $I_c$ can be calculated as:

$$ I_c = \frac{D_c - D_{min}}{N - D_{min}} $$

[Back to top]("#")

---

# Calculating $D_{min}$

## Partitions

$D_{min}$ depends on how $N$ individuals can be partitioned between $Z$ zones in 
the maze. 

In number theory and combinatorics, a partition of a positive integer $x$, also 
called an integer partition, is a way of writing $x$ as a sum of positive 
integers [@Andrews1998-pz; @Nathanson2000-dt; @Andrews2004-xu]. Two sums that 
differ only in the order of their summands are considered to be the same 
partition. The number of distinct partitions of $x$, without regard to the 
ordering of the terms, is defined as $P(x)$.

For example, 4 can be partitioned in five distinct ways; 

$$ 4;\quad 3 + 1;\quad 2 + 2;\quad 2 + 1 + 1;\quad 1 + 1 + 1 + 1 $$

Thus, the number of partitions $P(4) = 5$. As partitions are not order-dependent, 
$1 + 3$ is the same partition as $3 + 1$ for instance. It is conventional to 
write the parts of a partition in descending order.

In `R` [@R2016], the `partitions` package [@partitions2016] can be used to obtain
$P(x)$ (with the function `P`) as well as all the possible partitions of $x$
(with the function `parts`). For instance, for $x=4$:

```{r}
if (!require(partitions)) {
  install.packages("partitions")
  library(partitions)
}

x <- 4
P(x)
parts(x)
```

(with each column corresponding to a different partition of $x$).

[Back to top]("#")

---

## Partitions when $Z \geq N$

In the radial maze context, if the number of zones $Z$ is superior or equal to 
the number of individuals $N$, all the possible partitions of $N$ can be observed, 
and notably the more distributed partitions where each individual is alone in a 
zone). The number of partitions and the partitions themselves can therefore be 
obtained as described above. 

[Back to top]("#")

---

## Partitions when $Z < N$

If the number of zones $Z$ is inferior to the number of individuals $N$, then 
not all partitions can be observed. For instance, in the case of $N=4$ and $Z=3$,
it is impossible to observe the partition $1+1+1+1$. The total number of 
observable partitions in this case is therefore $P(4, 3) = 4$.

In `R`, the number of restricted partitions (i.e. the number of partitions that 
can be observed given $N$ and $Z$ in our case) can be calculated using the 
function `R` in the `partitions` package. For instance, for $N=4$ and $Z=3$:

```{r}
N <- 4
Z <- 3

R(Z, N, include.zero = TRUE)
```

And the restricted partitions can be obtained with the function `restrictedparts`
as follows: 

```{r}
restrictedparts(N, Z)
```

(with each column corresponding to a different partition of $N$ in $Z$ zones).

The Supplementary Table 1 below shows the number of restricted partitions for 
various values of $N$ and $Z$. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
if (!require(pander)) {
  install.packages("partitions")
  library(pander)
}

N <- 1:10
Z <- 1:10
res <- matrix(1, nrow = length(N), ncol = length(Z))

for (i in 1:length(N)) {
  for (j in 1:length(Z)) {
    if (Z[j] >= N[i]) {
      res[i, j] <- P(N[i])
    } else {
      res[i, j] <- R(Z[j], N[i], include.zero = TRUE)
    }
  }
}

res <- as.data.frame(res)

row.names(res) <- paste0("N=", N)
names(res) <- paste0("Z=", Z)

pander(res, style = "rmarkdown", split.table = Inf)
```

Table: **Supplementary Table 1:** Partition number as a function of the number 
of individuals $N$ and the number of zones $Z$. In our experiment, the partition 
number $P(10,7)$ is $38$.

[Back to top]("#")

---

## $D_{min}$ as a function of $N$ and $Z$

With `R`, $D_{min}$ can be easily calculated for a given $N$ and $Z$ as follows
(here $N=4$ and $Z=3$, as before):

```{r}
N <- 4
Z <- 3

partitions <- restrictedparts(N, Z)
Dc <- sqrt(apply(partitions ^ 2, 2, sum))
min(Dc)
```

The Supplementary Table 2 below shows the values of $D_{min}$ for various values
of $N$ and $Z$. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
N <- 1:10
Z <- 1:10
res <- matrix(1, nrow = length(N), ncol = length(Z))

for (i in 2:length(N)) {
  for (j in 2:length(Z)) {
    partitions <- restrictedparts(N[i], Z[j])
    Dc <- sqrt(apply(partitions ^ 2, 2, sum))
    res[i, j] <- min(Dc)
  }
}

int <- res %% 1 == 0
res <- as.data.frame(res)

res[!int] <- paste0("$\\sqrt{", res[!int] ^ 2, "}$")

row.names(res) <- paste0("N=", N)
names(res) <- paste0("Z=", Z)

pander(res, style = "rmarkdown", split.table = Inf)
```

Table: **Supplementary Table 2:** Minimal distance $D_{min}$ as a function of 
the number of individuals $N$ and the number of zones $Z$. When $Z \geq N$, 
$D_{min} = \sqrt{N}$. When $Z \leq N$, $D_{min}$ increases when $Z$ decreases. For 
the particular case where there is just one zone, $D_{max} = D_{min} = N$, in 
which case $I_c$ cannot be calculated. In our experiment, $D_{min} = 4$.

[Back to top]("#")

---

# Calculating $I_c$ with `R`

For a same number of individuals, $I_c$ values can be different if the number of 
zones varies, but $I_c$ values are always between 0 (worst group cohesion 
possible) and 1 (best group cohesion possible). When $Z \leq N$, $D_{min}$ 
increases as $Z$ decreases. In our experiment, with 7 zones and 10 fish, the 
partition number $P(10,7)$ is 38, and not 42 as in $P(10,10)$ or $P(10,12)$ for 
example; as a consequence and $D_c$ varies between $D_{max} = 10$ and $D_{min} = 4$. 

For convenience, we provide with our study a `R` package - called `projectRadial` 
containing a function - called `Ic` - for computing $I_c$ for any observed 
partitions of fish in radial arm mazes with any arbitrary number of arms. `Ic`
takes a single input, which should be a data frame containing the number of 
animals in each zone of the radial arm maze. The data frame should be organized 
as follows: each column must correspond to a zone of the radial arm maze and 
each row must correspond to a different observation.

An example of using `Ic` with actual data (10 fish in a radial maze with 6 arms 
and a central zone) is provided below: 

```{r}
# Install and load the 'projecRadial' package
if (!require(pacman)) {
  install.packages("pacman")
  library(partitions)
} 

p_load_gh("sjmgarnier/projectRadial")


# Calculate Ic with data provided in the 'projecRadial' package
data_path <- system.file("extdata", "sample_data.csv", package = "projectRadial")
dat <- read.csv(data_path)

dat$Ic <- Ic(dat[, 2:7])


# Plot
if (!require(ggplot2)) {
  install.packages("ggplot2")
  library(ggplot2)
}

ggplot(dat, aes(x = TimeInHours, y = Ic)) +
  geom_line() + 
  xlab("Time (hours)") + 
  ylab(bquote(I[c])) +
  theme_minimal(base_size = 18)
```

[Back to top]("#")

---

# References {-}

